Ecosystem size and complexity dictate riverine biodiversity

### 2020-11-16 manuscript/manuscript.Rmd
Edits: misc

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- name: Akira Terui
  email: hanabi0111@gmail.com
  affiliation: Department of Biology, University of North Carolina at Greensboro
  footnote: Corresponding Author
- name: Seoghyun Kim
  email: s_kim53@uncg.edu
  affiliation: Department of Biology, University of North Carolina at Greensboro
- name: Christy Dolph
  email: XXX
  affiliation: XXX
output: html_notebook
bibliography: reference.bib
csl: nature.csl
abstract: |
  This is the abstract.
  It consists of two paragraphs.
---

```{r setup, include=FALSE}
pacman::p_load(tidyverse, sf)
setwd(here::here("empirical"))

## read gis files
filename <- list.files(path = 'data_gis', full.names = T)
wsd_subset <- lapply(filename[str_detect(filename, "wsd_subset")], st_read)
point_subset <- lapply(filename[str_detect(filename, "point_subset")], st_read)
shape <- lapply(filename[str_detect(filename, "shape")], st_read)
shape[[1]] <- st_set_crs(shape[[1]], st_crs(wsd_subset[[1]]))

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area (the species-area relationship; SAR). In 1921, XXX formulated the SAR as a power-law $S = cA^z$ ($S$ is the number of species observed in a given geographic area $A$), the equation currently known as the Arrhenius species-area relationship. Since then, the spatial scaling of species richness has been observed in nearly all taxonomic groups, although the underlying mechanisms are still hotly debated. The SAR is general because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse communities due to increased habitat diversity, larger (meta)community size, and/or enhanced colonization dynamics. Importantly, the SAR provides the foundation for global conservation efforts. For example, conservation ecologists have used the SAR estimates to design marine and terrestrial protected areas, which currently encompass more than ~ km² globally.

However, many ecosystems possess a dimension of spatial complexity that cannot be described by area (i.e., the scale-invariant complexity). Such spatial complexity is evident in branching ecosystems, such as rivers, trees, and mountain ranges, to name just a few. Geomorphic or biological processes generate a pronounced self-similarity in complex branching structure so that the part and the whole look alike. Even though the branching structure is independent of spatial scale, it forms a physical template that determines habitat diversity and dispersal corridors for living organisms. Evidence suggests that ecological dynamics driven by branching structure cannot be predicted by scale-dependent theories; this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems. However, most research either has explored the consequences of branching complexity for local biodiversity patterns or relied on theoretical arguments with limited replicates of artificial branching networks. At present, no studies have integrated theoretical and empirical approaches to link branching complexity and $\gamma$ diversity (i. e., the total species richness in a metacommunity). This knowledge gap significantly limits our ability to guide biodiversity conservation in spatially complex ecosystems.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We constructed a general simulation model of metacommunity dynamics [@thompsonProcessbasedMetacommunityFramework2020], in which 50 competing species disperse and spatially interact in a random branching network. Simulated species with different niche in abiotic conditions (optimum and width) compete for resources with varied strengths. For simplicity, other ecological traits are assumed to be identical among species (e.g., dispersal probability). Here, branching networks are depicted as a spatial network of connected habitat patches (or a discretized river section) where mean environmental conditions at each habitat patch are determined by two factors: (1) environments of headwaters (the most upstream habitat patch) and (2) local environmental noises. Environmental values at headwaters are produced randomly and propagate downstream with local environmental noises (i.e., the spatial autoregressive process with  a white noise). These values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our simulation resembles ecological processes of how branching river networks create diverse habitats in a metacommunity. We prepared 16 simulation scenarios with varied ecological (dispersal and competition strength) and landscape contexts (environmental variation at headwaters and the degree of local environmental noises). Under each scenario, we simulated metacommunity dynamics in 1000 branching networks that have different ecosystem size (the number of habitat patches in a network) and complexity (branching probability) to examine their effects on $\gamma$ diversity.

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under realistic landscape contexts, where the environmental variation at headwaters ($\sigma_h$) is greater than the degree of local environmental noises ($\sigma_l$) [@petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. The relationships had a characteristic of power-law (i.e., linear in a log-log scale) and were consistent under various ecological scenarios: the strength of competition and dispersal probability did not change the form of the relationships between $\gamma$ diversity and ecosystem properties. The effect of ecosystem size on $\gamma$ diversity was largely driven by the spatial turnover of species composition among habitat patches ($\beta$ diversity) regardless of ecological scenarios. $\beta$ diversity increased significantly with increasing ecosystem size, while local diversity (average species richness in a single habitat patch; $\alpha$ diversity) remained almost unchanged.

In contrast, the relative importance of $\alpha$ and $\beta$ diversity in the complexity-$\gamma$ diversity relationship was dependent on the dispersal probability of constituent species. When dispersal probability of species was low, the positive effect of branching complexity on $\gamma$ diversity was driven mainly by increased $\beta$ diversity, reflecting significant spatial turnover of species composition. However, with higher dispersal probability, the relationship between $\beta$ diversity and branching complexity became a hump-shaped form. A clear increase in $\alpha$ diversity compensated for the slight decrease in $\beta$ diversity at high branching probabilities, maintaining the positive complexity-$\gamma$ diversity relationship. This likely occurred through the increased connectivity of highly branched networks. The average network distance between a pair of habitat patches decreases with increasing branching probability because highly branched networks have a compressed form compared to less branched networks. Therefore, increased dispersal readily homogenizes community composition while increasing $\alpha$ diversity through enhanced immigration from suitable habitat patches (i.e., mass effects). This result highlights how different mechanisms can generate an apparently similar pattern in the complexity-$\gamma$ diversity relationship.

It is important to note that the effect of branching complexity diminished when the environmental variation at headwaters is smaller or equal to the degree of local environmental noises ($\sigma_h \le \sigma_l$). When local environmental noises are very large relative to environmental variation among headwaters, the net habitat diversity should be correlated with ecosystem size but not with branching complexity. This simulation setup decouples the intimate relationship between branching structure and habitat diversity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to natural river networks because streams originating from different sources are known to show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes. Yet, this result illustrates the crucial role of habitat diversity produced by branching structure, rather than branching structure itself, in the emerging complexity-$\gamma$ diversity relationship.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyn Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior]
output:
  html_notebook: default
abstract: |
  |This is the abstract.
bibliography: reference.bib
csl: nature.csl
---

```{r setup, include=FALSE}

pacman::p_load(tidyverse, sf)

## read gis files
filename <- list.files(path = normalizePath('../empirical/data_gis'), full.names = T)
wsd_subset <- lapply(filename[str_detect(filename, "wsd_subset")], st_read)
point_subset <- lapply(filename[str_detect(filename, "point_subset")], st_read)
shape <- lapply(filename[str_detect(filename, "shape")], st_read)
shape[[1]] <- st_set_crs(shape[[1]], st_crs(wsd_subset[[1]]))

## read theory file
dat_sim <- read_csv(normalizePath('../theory/result/result_sim2020-11-10.csv'))

n_scenario <- dat_sim %>% distinct(n_species,
                     mean_env_source,
                     sd_env_source,
                     rho,
                     asymmetry_factor,
                     min_optim,
                     max_optim,
                     sd_env_lon,
                     theta,
                     K0,
                     sd_env,
                     spatial_env_cor,
                     phi,
                     min_niche_width,
                     max_niche_width,
                     niche_cost,
                     p_dispersal,
                     min_alpha,
                     max_alpha) %>% 
  nrow()

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area (the species-area relationship; SAR). In 1921, XXX formulated the SAR as a power-law $S = cA^z$ ($S$ is the number of species observed in a given geographic area $A$), which is currently called as the Arrhenius species-area relationship. Since then, the spatial scaling of species richness has been observed in nearly all taxonomic groups, although the underlying mechanisms are still hotly debated. The SAR is general because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity, larger metacommunity size, and/or enhanced colonization dynamics. Importantly, the SAR provides the foundation for global conservation efforts. For example, conservation ecologists have used the SAR estimates to design marine and terrestrial protected areas, which currently encompass more than ~ km² globally.

Many ecosystems, however, possess complex spatial structure that cannot be represented by area, a dimension known as the scale-invariant complexity. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike. Even though the branching structure is independent of spatial scale, it forms a physical template that determines habitat diversity and dispersal corridors for living organisms. Evidence suggests that scale-dependent theories cannot predict ecological dynamics driven by branching structure, and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research either has explored the consequences of branching complexity for local biodiversity patterns or relied on theoretical arguments with limited replicates of artificial branching networks. At present, no studies have investigated how ecosystem size and complexity jointly regulate species diversity at the metacommunity level (i. e., $\gamma$ diversity). This knowledge gap significantly limits our ability to guide biodiversity conservation in spatially complex ecosystems.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We constructed a general simulation model of metacommunity dynamics [@thompsonProcessbasedMetacommunityFramework2020], in which 50 competing species disperse and spatially interact in a random branching network. Simulated species with different niche in abiotic conditions (optimum and width) compete for resources with varied strengths. For simplicity, other ecological traits, such as dispersal probability, are assumed to be identical among species. Here, branching networks are depicted as a spatial network of connected habitat patches (or a discretized river section; see **Figure 1**) where mean environmental conditions at each habitat patch are determined by two factors: (1) the headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at the headwaters are drawn randomly from a normal distribution and propagates downstream with local environmental noise (i.e., the spatial autoregressive process with white noise). These values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our simulation resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We prepared `r n_scenario` simulation scenarios with varied ecological (dispersal and competition strength) and landscape contexts (the environmental variation at headwaters and the degree of local environmental noise). Under each scenario, we simulated metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks that have different ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`) to examine their effects on $\gamma$ diversity.

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape context (**Figures 2 and 3**), where the environmental variation at headwaters ($\sigma_h$) is greater than the degree of local environmental noise ($\sigma_l$) [@petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. The relationships had a characteristic of power-law (i.e., linear in a log-log scale) and were consistent under various ecological scenarios. The strength of competition and dispersal processes (dispersal distance and probability) did not change the form of the relationships between $\gamma$ diversity and ecosystem properties. Hence, the ecosystem size and complexity are both expected to increase $\gamma$ diversity regardless of ecological scenarios.

Dispersal processes, however, affected underlying mechanisms that produce the positive effects of ecosystem size and complexity on $\gamma$ diversity. We observed a greater contribution of $\beta$ diversity, which we defined as $\frac{\alpha}{\gamma}$, to increased $\gamma$ diversity in larger and more complex ecosystems when dispersal limitation exists (i.e., species travel short distances). This pattern reflects significant spatial turnover of species composition over the branching network. In contrast, once the dispersal limitation is relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. This is understandable because increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects). These patterns were consistent across different levels of emigration probabilities (**cite SI figure**). Our theory, therefore, highlights how the apparently similar patterns in $\gamma$ diversity emerges through different ecological pathways.

Influences of ecosystem size and complexity, however, differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of the net habitat diversity in the ecosystem (environmental variation among headwaters + local environmental noise), though the slopes were steeper when environmental variation was greater. This is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity. In contrast, the effect of branching complexity depended strongly on the environmental variation at headwaters ($\sigma_h$; see **XXX**). If environmental variation among headwaters is small, there is no foundation for branching complexity to produce diverse environmental conditions over the landscape. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat diversity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinctiveness in a variety of environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes. Yet, this result illustrates the crucial role of habitat diversity produced by branching structure in the emerging complexity-$\gamma$ diversity relationship.
